---
title: "Blood DNA Methylation Signature for Incident Dementia: Evidence from Longitudinal Cohorts"
subtitle: "Cox Regression on Single CpGs --- FHS"
author: 
  - Wei Zhang^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    highlight: pygments
    theme: yeti
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
# Initial settings
knitr::opts_chunk$set(echo = TRUE, eval = T, warning = F, message = F)

# Create dir
cohort <- "Framingham"
dir.base <- "../../.."
dir.data <- file.path(dir.base,"../DATASETS/", cohort)
dir.data.raw <- file.path(dir.data, "DNAm/raw/EPIC") 
dir.data.processed <- file.path(dir.data, "DNAm/processed/EPIC") 
dir.data.aux <- file.path(dir.base,"../DATASETS/Aux_Sync/") 
dir.results <- file.path(dir.base, "analysis_results", cohort, "EPIC_EXAM9")
dir.results.data <- file.path(dir.base, "datasets", cohort) 
dir.results.batch <- file.path(dir.base, "analysis_results/BEclear")
for(p in grep("dir.",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
```

```{r message = F}     
source("../../utility/cpg_test.R")
source("../../utility/annotation_and_bacon.R")
```

# Load data

```{r}
framingham_se <- readRDS(
  file.path(dir.data.processed, "FHS_OffSpring_EPIC_rm_dem_nw_se.RDS")
)
celltype <- c("B", "NK", "CD4T", "Mono", "Gran")
betaMat <- assay(framingham_se)
pheno_df <- colData(framingham_se) %>% data.frame()
pheno_df <- pheno_df %>% 
  mutate(across(all_of(celltype), ~ .x*100))
```

# Check distribution

```{r}
# age
mean(pheno_df$Age_at_collection)
sd(pheno_df$Age_at_collection)
# follow up time
mean(pheno_df$dem_followup_time)/365
sd(pheno_df$dem_followup_time)/365
# sex
pheno_df %>% 
  summarise(n = n(), 
            freq = n/nrow(pheno_df) * 100,
            .by = sex)
# education
mean(pheno_df$EDUC_YEAR, na.rm = T)
sd(pheno_df$EDUC_YEAR, na.rm = T)
# smoking
pheno_df %>% 
  summarise(n = n(), 
            freq = n/nrow(pheno_df) * 100,
            .by = smoking_status)
pheno_df <- pheno_df %>% 
  mutate(smoking_hist = ifelse(is.na(smoking_status), NA, ifelse(
         smoking_status %in% c(1,2), 1, 0)))
```

```{r}
plyr::ldply(
  c("dem_followup_time", "Age_at_collection",  "sex", "EDUC_YEAR", "smoking_hist"),
  .fun = function(var) {
    if(!var %in% c("sex", "smoking_hist")) {
      fo <- as.formula(paste0(var, "~ DEM_STATUS"))
      df <- pheno_df %>% rstatix::wilcox_test(fo) %>% 
        dplyr::select(var = .y., p)
    } else {
      df <- pheno_df %>% rstatix::freq_table(var, "DEM_STATUS")
      mat <- as.table(matrix(df$n, nrow = 2, ncol = 2))
      dimnames(mat) <- list(DEM_STATUS = df$DEM_STATUS[1:2],
                            GROUP = df[[var]][c(1,3)])
      df <- rstatix::fisher_test(mat) %>%
        dplyr::mutate(var = var) %>%
        dplyr::select(var, p)
    }
    df
  }
)
```

## Correlation between PC and covariates

```{r}
# compute sds for each row
sds <- matrixStats::rowSds(betaMat)
sdsSorted <- order(sds, decreasing = TRUE)

# order by most variable probes on top
beta <- betaMat[sdsSorted ,]
pc <- irlba::irlba(t(beta[1:50000,]))
pl <- plyr::llply(
   c("dem_followup_time", "Age_at_collection",  "sex", "EDUC_YEAR", "smoking_hist"),
  .fun = function(var) {
    if(!var %in% c("sex", "smoking_hist")) {
      if(var == "dem_followup_time") {
        xlab <- "Follow up period in years"
        pheno_df$dem_followup_time <- pheno_df$dem_followup_time/365
      } 
      if(var == "Age_at_collection") {
        xlab <- "Age"
      }
      if(var == "EDUC_YEAR") {
        xlab <- "Education"
      }
      if(var != "EDUC_YEAR") {
        
        ggpubr::ggscatter(
          data = data.frame(pheno_df, PC = pc$u[,1]),
          x = var,
          y = "PC",
          cor.coef = T,
          add = "reg.line",
          ylab = "First PC of FHS beta values",
          xlab = xlab,
          add.params = list(color = "navy"),
          cor.method = "spearman",
          cor.coeff.args = list(label.y.npc = "top")
        )
      } else {
        df <- data.frame(pheno_df, PC = pc$u[,1])
        df <- df[!is.na(df[[var]]),]
        ggpubr::ggboxplot(
          data = df,
          x = var,
          y = "PC",
          fill = "grey",
          palette = "npg",
          ylab = "First PC of FHS beta values",
          xlab = xlab
        ) + ggpubr::stat_compare_means(method = "kruskal.test", label.y.npc = "bottom", label.x.npc = 0.3) + 
        guides(fill = "none")
      }
    } else {
      if (var == "smoking_hist") xlab <- "Smoking History"
      if (var == "sex") xlab <- "Gender"
      df <- data.frame(pheno_df, PC = pc$u[,1])
      df <- df[!is.na(df[[var]]),]
      ggpubr::ggboxplot(
        data = df,
        x = var,
        y = "PC",
        fill = var,
        palette = "npg",
        ylab = "First PC of FHS beta values",
        xlab = xlab
      ) + ggpubr::stat_compare_means(method = "wilcox.test", label.y.npc = "bottom") + 
        guides(fill = "none")
    }
  }
)
ggpubr::ggarrange(plotlist = pl, nrow = 3, ncol = 2)
ggsave(
  filename = file.path(dir.results, "FHS_beta_pc_vs_covariates.pdf"),
  height = 12,
  width = 10
)
```

# Main analysis

## Fit model 

```{r eval = F}
results <- cox_coef(
  beta = betaMat,
  pheno = pheno_df,
  time_var = "dem_followup_time",
  event_var = "DEM_STATUS",
  adjust_var = c("Age_at_collection", "sex", celltype),
  scale = T,
  fdr_method = "fdr",
  save = F
)
```

```{r eval = F}
write_csv(
  results, 
  file.path(dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_results.csv")
)
```

## Add annotation

```{r eval = F}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
# Load from annotation_and_bacon.R
results_anno <- annotate_results(results,
                                 array = "EPIC",
                                 dir.data.aux = dir.data.aux,
                                 dir.save = dir.results,
                                 prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes")
```

```{r}
results_anno <- read_csv(
  file.path(dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_annotated_results.csv")
)
```

```{r}
bacon_results <- bacon_adj(
  results_anno,
  est_var = "coef",
  z_var = "z",
  std_var = "se_coef"
)
```

```{r eval = F}
writexl::write_xlsx(
  bacon_results$inflation.stat,
  file.path(dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_inflation_stats.xlsx")
)
```

```{r eval = F}
write_csv(
  bacon_results$data.with.inflation,
  file.path(dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_bacon_results.csv")
)
```

# Check with pedno

```{r message = F}
library(lme4)
```

```{r}
bacon_results <- read_csv(
  file.path(dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_bacon_results.csv")
)
bacon_results_sig <- bacon_results %>% 
  filter(pValue.bacon < 1e-05)
```

```{r warning = F, message = F}
betaMat <- assay(framingham_se)
pheno <- colData(framingham_se) %>% data.frame()
M <- minfi::logit2(betaMat)
M <- M[bacon_results_sig$cpg,]

results_mm <- plyr::adply(
    M,
    1,
    .fun = function(mvalue){
      mod = lmer(mvalue ~ (1|pedno), data.frame(mvalue = mvalue, pheno))
      vc <- VarCorr(mod) %>% as.data.frame()
      data.frame(ICC = vc$vcov[1]/(vc$vcov[1] + vc$vcov[2]))
    },.id = "cpg"
)

write_csv(
  results_mm,
  file.path(dir.results, "ICC_pedno_cpg_1e05.csv")
)
```

# Sensitivity analysis 1: smoking

```{r eval = F}
results <- cox_coef(
  beta = betaMat,
  pheno = pheno_df,
  time_var = "dem_followup_time",
  event_var = "DEM_STATUS",
  adjust_var = c("Age_at_collection", "sex", celltype, "smoking_hist"),
  scale = T,
  fdr_method = "fdr",
  save = T,
  dir.save = dir.results,
  prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_smokhist",
  cores = 20
)
```

## Add annotation

```{r eval = F}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
# Load from annotation_and_bacon.R
results_anno <- annotate_results(results,
                                 array = "EPIC",
                                 dir.data.aux = dir.data.aux,
                                 dir.save = dir.results,
                                 prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_smokhist")
```

```{r}
results_anno <- read_csv(
  file.path(dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_smokhist_annotated_results.csv")
)
```

```{r}
bacon_results <- bacon_adj(
  results_anno,
  est_var = "coef",
  z_var = "z",
  std_var = "se_coef",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_smokhist"
)
```

# Sensitivity analysis 2: exclude high-risk individual

```{r}
followup_qt <- quantile(pheno_df$dem_followup_time)
tau_qt <- quantile(as.numeric(pheno_df$TAU), na.rm = T)
pheno_df <- pheno_df %>% 
  filter((DEM_STATUS == 0 & dem_followup_time > followup_qt[2]) | (DEM_STATUS == 0 & TAU < tau_qt[4]) | DEM_STATUS == 1)
beta <- betaMat[,pheno_df$sample]
```

```{r eval = F}
# This function is load from cpg_test.R
results <- cox_coef(
  beta = beta,
  pheno = pheno_df,
  time_var = "dem_followup_time",
  event_var = "DEM_STATUS",
  adjust_var = c("Age_at_collection", "sex", celltype),
  scale = T,
  fdr_method = "fdr",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_rm_highrisk",
  cores = 20
)
```

## Add annotation

```{r eval = F}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
# This function is load from annotation_and_bacon.R
results_anno <- annotate_results(results,
                                 array = "EPIC",
                                 dir.data.aux = dir.data.aux,
                                 dir.save = dir.results,
                                 prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_rm_highrisk")
```

## Bacon correction

```{r message = F, echo = F}
results_anno <- read_csv(file.path(
  dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_rm_highrisk_annotated_results.csv"
))
```

```{r}
# This function is load from annotation_and_bacon.R
results_bacon <- bacon_adj(
  data = results_anno,
  est_var = "coef",
  z_var = "z",
  std_var = "se_coef",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_rm_highrisk")
```

# Sensitivity analysis 3: AD in Dementia

```{r eval = F}
# This function is load from cpg_test.R
results <- cox_coef(
  beta = betaMat,
  pheno = pheno_df,
  time_var = "dem_followup_time",
  event_var = "AD_STATUS",
  adjust_var = c("Age_at_collection", "sex", celltype),
  scale = T,
  fdr_method = "fdr",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_AD",
  cores = 30
)
```

## Add annotation

```{r eval = F}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
# This function is load from annotation_and_bacon.R
results_anno <- annotate_results(results,
                                 array = "EPIC",
                                 dir.data.aux = dir.data.aux,
                                 dir.save = dir.results,
                                 prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_AD")
```

## Bacon correction

```{r message = F, echo = F}
results_anno <- read_csv(file.path(
  dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_AD_annotated_results.csv"
))
```

```{r}
# This function is load from annotation_and_bacon.R
results_bacon <- bacon_adj(
  data = results_anno,
  est_var = "coef",
  z_var = "z",
  std_var = "se_coef",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_AD")
```

# Compare

```{r}
results1 <- read_csv(
  file.path(dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_bacon_results.csv")
)
results2 <- read_csv(
  file.path(dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_smokhist_bacon_correction.csv")
)
results3 <- read_csv(
  file.path(dir.results, "Framingham_EPIC_beta_single_cpg_cox_all_celltypes_rm_highrisk_bacon_correction.csv")
)
```

```{r}
ls <- list(
  Original = results1 %>% filter(pValue.bacon < 1e-05) %>% pull(cpg),
  Sensitivity = results2 %>% filter(pValue.bacon < 1e-05) %>% pull(cpg),
  "Remove high-risk individuals" = results3 %>% filter(pValue.bacon < 1e-05) %>% pull(cpg)
)
ggvenn::ggvenn(ls)
```

# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session}
sessionInfo()
```
</details>
