---
title: "Blood DNA Methylation Signature for Incident Dementia: Evidence from Longitudinal Cohorts"
subtitle: "Cox Regression on Single CpGs --- ADNI"
author: 
  - Wei Zhang^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    highlight: pygments
    theme: yeti
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
cohort <- "ADNI"
dir.base <- "../../.."
dir.results <- file.path(dir.base, "analysis_results", cohort, "cox")
dir.results.data <- file.path(dir.base, "datasets", cohort, "BEclear") 
dir.data.aux <- file.path(dir.base,"../DATASETS/Aux_Sync/") 
```

# Load data and auxilary functions

```{r message = F, warning = F}
source(file.path(dir.base, "/code/utility/cpg_test.R"))
source(file.path(dir.base, "/code/utility/annotation_and_bacon.R"))
```

```{r}
adni_se_corrected <- readRDS(
  file.path(dir.results.data, "ADNI_BEclear_se.rds")
)
```

```{r}
betaMat <- assay(adni_se_corrected)
pheno_df <- colData(adni_se_corrected) %>% 
  as.data.frame() 

# pheno_df <- write_csv(
#   pheno_df,
#   file.path(dir.base, "../DATASETS", cohort, "Phenotype/processed/all_sample/ADNI_cn_baseline_first_visit_pheno.csv")
# )
```

# Check distribution

```{r}
# age
mean(pheno_df$age_at_visit)
sd(pheno_df$age_at_visit)
# follow up time
mean(pheno_df$SurvDays)/365
sd(pheno_df$SurvDays)/365
# sex
pheno_df %>% 
  summarise(n = n(), 
            freq = n/nrow(pheno_df) * 100,
            .by = PTGENDER)
# education
mean(pheno_df$PTEDUCAT, na.rm = T)
sd(pheno_df$PTEDUCAT, na.rm = T)
# smoking
medhist <- read_csv(
  file.path(dir.base, "../DATASETS", cohort, "Phenotype/raw/MEDHIST.csv")
)
medhist <- medhist %>%
  dplyr::select(RID, MH16SMOK) %>% 
  group_by(RID) %>%
  slice_max(MH16SMOK, with_ties = F) %>% # For each participant (RID), if any record indicates a history of smoking (MH16SMOK = 1), that row is retained.
  ungroup()
pheno_df <- left_join(pheno_df, medhist)
```

## Tests

```{r}
plyr::ldply(
  c("SurvDays", "age_at_visit",  "PTGENDER", "PTEDUCAT", "MH16SMOK"),
  .fun = function(var) {
    if(!var %in% c("PTGENDER", "MH16SMOK")) {
      fo <- as.formula(paste0(var, "~ DEM_STATUS"))
      df <- pheno_df %>% rstatix::wilcox_test(fo) %>% 
        dplyr::select(var = .y., p)
    } else {
      df <- pheno_df %>% rstatix::freq_table(var, "DEM_STATUS")
      mat <- as.table(matrix(df$n, nrow = 2, ncol = 2))
      dimnames(mat) <- list(DEM_STATUS = df$DEM_STATUS[1:2],
                            GROUP = df[[var]][c(1,3)])
      df <- rstatix::fisher_test(mat) %>%
        dplyr::mutate(var = var) %>%
        dplyr::select(var, p)
    }
    df
  }
)
```

```{r}
celltype <- c("B", "NK", "CD4T", "Mono", "Gran")
adni_se_corrected$PTGENDER <- as.factor(adni_se_corrected$PTGENDER)
betaMat <- assay(adni_se_corrected)
pheno <- colData(adni_se_corrected) %>% data.frame()
# Multiply 100 for celltype proportion
pheno <- pheno %>% 
  mutate(across(all_of(celltype), ~ .x*100))
```

```{r}
pca_cell <- prcomp(
  pheno[,celltype]
)
```

## Correlation between PC and covariates

```{r}
# compute sds for each row
sds <- matrixStats::rowSds(betaMat)
sdsSorted <- order(sds, decreasing = TRUE)

# order by most variable probes on top
beta <- betaMat[sdsSorted ,]
pc <- irlba::irlba(t(beta[1:50000,]))
pl <- plyr::llply(
  c("SurvDays", "age_at_visit",  "PTGENDER", "PTEDUCAT", "MH16SMOK"),
  .fun = function(var) {
    print(var)
    if(!var %in% c("PTGENDER", "MH16SMOK")) {
      if(var == "SurvDays") {
        xlab <- "Follow up period in years"
        pheno_df$SurvDays <- pheno_df$SurvDays/365
      } 
      if(var == "age_at_visit") {
        xlab <- "Age"
      }
      if(var == "PTEDUCAT") {
        xlab <- "Education"
      }
      if(var != "PTEDUCAT") {
        
        ggpubr::ggscatter(
          data = data.frame(pheno_df, PC = pc$u[,1]),
          x = var,
          y = "PC",
          cor.coef = T,
          add = "reg.line",
          ylab = "First PC of ADNI beta values",
          xlab = xlab,
          add.params = list(color = "navy"),
          cor.method = "spearman",
          cor.coeff.args = list(label.y.npc = "top")
        )
      } else {
        ggpubr::ggboxplot(
          data = data.frame(pheno_df, PC = pc$u[,1]),
          x = var,
          y = "PC",
          fill = "grey",
          palette = "npg",
          ylab = "First PC of ADNI beta values",
          xlab = xlab
        ) + ggpubr::stat_compare_means(method = "kruskal.test", label.y.npc = "bottom") + 
        guides(fill = "none")
      }
      
    } else {
      if (var == "MH16SMOK") xlab <- "Smoking History"
      if (var == "PTGENDER") xlab <- "Gender"
      ggpubr::ggboxplot(
        data = data.frame(pheno_df, PC = pc$u[,1]),
        x = var,
        y = "PC",
        fill = var,
        palette = "npg",
        ylab = "First PC of ADNI beta values",
        xlab = xlab
      ) + ggpubr::stat_compare_means(method = "wilcox.test", label.y.npc = "bottom") + 
        guides(fill = "none")
    }
  }
)
ggpubr::ggarrange(plotlist = pl, nrow = 3, ncol = 2)
ggsave(
  filename = file.path(dir.results, "ADNI_beta_pc_vs_covariates.pdf"),
  height = 12,
  width = 10
)
```


# Main analysis

## Analysis with firt two PCs of cell type proportions

```{r eval = F}
# This function is load from cpg_test.R
results <- cox_coef(
  beta = betaMat,
  pheno = data.frame(pheno, pca_cell$x),
  time_var = "SurvDays",
  event_var = "DEM_STATUS",
  adjust_var = c("age_at_visit", "PTGENDER", "PC1", "PC2"),
  scale = T,
  fdr_method = "fdr",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "ADNI_adjust_celltypes_first_2pc",
  cores = 20
)
```

```{r message = F, echo = F}
results <- read_csv(file.path(
  dir.results, "ADNI_adjust_celltypes_first_2pc_single_cpg_cox_results.csv"
))
```

## Add annotation

```{r eval = F}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
# This function is load from annotation_and_bacon.R
results_anno <- annotate_results(results,
                                 array = "EPIC",
                                 dir.data.aux = dir.data.aux,
                                 dir.save = dir.results,
                                 prefix = "ADNI_adjust_celltypes_first_2pc_single_cpg_cox")
```

## Bacon correction

```{r message = F, echo = F}
results_anno <- read_csv(file.path(
  dir.results, "ADNI_adjust_celltypes_first_2pc_single_cpg_cox_annotated_results.csv"
))
```

```{r}
# This function is load from annotation_and_bacon.R
results_bacon <- bacon_adj(
  data = results_anno,
  est_var = "coef",
  z_var = "z",
  std_var = "se_coef",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "ADNI_adjust_celltypes_first_2pc_single_cpg_cox")
```

# Sensitivity analysis 1: smoking

```{r eval = F}
# This function is load from cpg_test.R
results <- cox_coef(
  beta = betaMat,
  pheno = data.frame(pheno, pca_cell$x),
  time_var = "SurvDays",
  event_var = "DEM_STATUS",
  adjust_var = c("age_at_visit", "PTGENDER", "PC1", "PC2", "MH16SMOK"),
  scale = T,
  fdr_method = "fdr",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "ADNI_adjust_celltypes_first_2pc_smokhist",
  cores = 20
)
```

## Add annotation

```{r eval = F}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
# This function is load from annotation_and_bacon.R
results_anno <- annotate_results(results,
                                 array = "EPIC",
                                 dir.data.aux = dir.data.aux,
                                 dir.save = dir.results,
                                 prefix = "ADNI_adjust_celltypes_first_2pc_smokhist_single_cpg_cox")
```

## Bacon correction

```{r message = F, echo = F}
results_anno <- read_csv(file.path(
  dir.results, "ADNI_adjust_celltypes_first_2pc_smokhist_single_cpg_cox_annotated_results.csv"
))
```

```{r}
# This function is load from annotation_and_bacon.R
results_bacon <- bacon_adj(
  data = results_anno,
  est_var = "coef",
  z_var = "z",
  std_var = "se_coef",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "ADNI_adjust_celltypes_first_2pc_smokhist_single_cpg_cox")
```

# Sensitivity analysis 2: exclude high-risk individual

```{r}
followup_qt <- quantile(pheno_df$SurvDays)
tau_qt <- quantile(as.numeric(pheno_df$TAU), na.rm = T)
pheno_df <- pheno_df %>% 
  filter((DEM_STATUS == 0 & SurvDays > followup_qt[2]) | (DEM_STATUS == 0 & TAU < tau_qt[4]) | DEM_STATUS == 1)
beta <- betaMat[,pheno_df$barcodes]
```

```{r eval = F}
# This function is load from cpg_test.R
results <- cox_coef(
  beta = beta,
  pheno = pheno_df,
  time_var = "SurvDays",
  event_var = "DEM_STATUS",
  adjust_var = c("age_at_visit", "PTGENDER", "PC1", "PC2"),
  scale = T,
  fdr_method = "fdr",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "ADNI_adjust_celltypes_first_2pc_rm_highrisk",
  cores = 20
)
```

## Add annotation

```{r eval = F}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
# This function is load from annotation_and_bacon.R
results_anno <- annotate_results(results,
                                 array = "EPIC",
                                 dir.data.aux = dir.data.aux,
                                 dir.save = dir.results,
                                 prefix = "ADNI_adjust_celltypes_first_2pc_rm_highrisk_single_cpg_cox")
```

## Bacon correction

```{r message = F, echo = F}
results_anno <- read_csv(file.path(
  dir.results, "ADNI_adjust_celltypes_first_2pc_rm_highrisk_single_cpg_cox_annotated_results.csv"
))
```

```{r}
# This function is load from annotation_and_bacon.R
results_bacon <- bacon_adj(
  data = results_anno,
  est_var = "coef",
  z_var = "z",
  std_var = "se_coef",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "ADNI_adjust_celltypes_first_2pc_rm_highrisk_single_cpg_cox")
```

# Sensitivity analysis 3: AD in Dementia

```{r}
dxsum <- read_csv(
  file.path(dir.data.aux, "../ADNI/Phenotype/raw/DXSUM_PDXCONV_ADNIALL_downloaded_12-11-2022.csv")
)
dxsum <- dxsum %>% 
  dplyr::select(RID, EXAMDATE, DIAGNOSIS, DXDDUE) %>% 
  group_by(RID) %>% 
  slice_max(as.Date(EXAMDATE))
dxsum$EXAMDATE <- NULL
pheno_df <- left_join(
  pheno, dxsum
)
table(pheno_df$DXDDUE)
```

```{r}
pheno_df$AD_STATUS <- ifelse(
  pheno_df$DXDDUE == 1, 1, 0
)
pheno_df$AD_STATUS[is.na(pheno_df$AD_STATUS)] <- 0
pheno_df <- data.frame(pheno_df, pca_cell$x)
```

```{r eval = F}
# This function is load from cpg_test.R
results <- cox_coef(
  beta = betaMat,
  pheno = pheno_df,
  time_var = "SurvDays",
  event_var = "AD_STATUS",
  adjust_var = c("age_at_visit", "PTGENDER", "PC1", "PC2"),
  scale = T,
  fdr_method = "fdr",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "ADNI_adjust_celltypes_first_2pc_AD",
  cores = 30
)
```

## Add annotation

```{r eval = F}
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
# This function is load from annotation_and_bacon.R
results_anno <- annotate_results(results,
                                 array = "EPIC",
                                 dir.data.aux = dir.data.aux,
                                 dir.save = dir.results,
                                 prefix = "ADNI_adjust_celltypes_first_2pc_AD_single_cpg_cox")
```

## Bacon correction

```{r message = F, echo = F}
results_anno <- read_csv(file.path(
  dir.results, "ADNI_adjust_celltypes_first_2pc_rm_highrisk_single_cpg_cox_annotated_results.csv"
))
```

```{r}
# This function is load from annotation_and_bacon.R
results_bacon <- bacon_adj(
  data = results_anno,
  est_var = "coef",
  z_var = "z",
  std_var = "se_coef",
  save = T,
  dir.save = file.path(dir.results),
  prefix = "ADNI_adjust_celltypes_first_2pc_AD_single_cpg_cox")
```

# Compare

```{r}
results1 <- read_csv(
  file.path(dir.results, "ADNI_adjust_celltypes_first_2pc_single_cpg_cox_bacon_correction.csv")
)
results2 <- read_csv(
  file.path(dir.results, "ADNI_adjust_celltypes_first_2pc_smokhist_single_cpg_cox_bacon_correction.csv")
)
results3 <- read_csv(
  file.path(dir.results, "ADNI_adjust_celltypes_first_2pc_rm_highrisk_single_cpg_cox_bacon_correction.csv")
)
results4 <- read_csv(
  file.path(dir.results, "ADNI_adjust_celltypes_first_2pc_AD_single_cpg_cox_bacon_correction.csv")
)
```

```{r fig.width = 8, fig.height = 8}
ls <- list(
  "Remove high-risk\nindividuals" = results3 %>% filter(pValue.bacon < 1e-05) %>% pull(cpg),
  Original = results1 %>% filter(pValue.bacon < 1e-05) %>% pull(cpg),
  Sensitivity = results2 %>% filter(pValue.bacon < 1e-05) %>% pull(cpg),
  "AD cox regression" = results4 %>% filter(pValue.bacon < 1e-05) %>% pull(cpg)
)
ggvenn::ggvenn(ls, set_name_size = 3)
```

```{r}
results <- read_csv(
  file.path(dir.results, "../../meta_analysis/meta_analysis_FHS9_ADNI_sig_results_annotated.csv")
)
sig.cpg <- results1$cpg[results$fdr < .05]
df <- results %>% filter(fdr < .05)
df2 <- results4 %>% 
  dplyr::select(cpg, Estimate.bacon, StdErr.bacon, pValue.bacon, fdr.bacon) %>% 
  rename_with( ~paste0("AD_", .x), contains("bacon"))
df <- left_join(df, df2)
df

write_csv(
  df,
  file.path(dir.results, "ADNI_sig_cpg_dem_vs_ad.csv")
)
```

# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session}
sessionInfo()
```
</details>
